---
title: "edgestream: streaming graph analytics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{edgestream: streaming graph analytics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Most graph tools assume a static graph: load everything, run an algorithm,
done. `edgestream` instead keeps its metrics up to date on every edge, so you
can query a graph that never stops growing — connected components, triangle
counts, degrees and clustering are maintained incrementally, and centrality
(betweenness, PageRank) is computed on demand.

Node IDs are non-negative integers, **0-indexed**. The graph object is an
external pointer to a C++ structure; pass it as the first argument to every
function.

## A streaming session

The canonical use case: transactions arrive between accounts and we want live
answers about connectivity and clustering.

```{r}
library(edgestream)

G   <- stream_graph()
idx <- node_index()   # maps arbitrary keys to integer node IDs

transactions <- list(
  c("acc_001", "acc_017"),
  c("acc_017", "acc_999"),
  c("acc_999", "acc_001"),   # closes a triangle
  c("acc_500", "acc_777")
)
for (tx in transactions) {
  add_edge(G, node_id(idx, tx[1]), node_id(idx, tx[2]))
}

same_component(G, get_id(idx, "acc_001"), get_id(idx, "acc_999"))
n_components(G)
triangle_count(G)
clustering_coefficient(G, get_id(idx, "acc_001"))
```

The scalar counts above are maintained O(1) reads and component lookup is
amortised O(alpha(n)). Methods returning vectors scale with their output; the
first component query after removals rebuilds DSU once.

## Weighted edges

Pass `weighted = TRUE` to store a positive weight per edge. Weights feed
`strength()`, PageRank and (Dijkstra-based) betweenness.

```{r}
W <- stream_graph(weighted = TRUE)
add_edge(W, 0L, 1L, w = 2.5)
add_edge(W, 1L, 2L, w = 1.0)
edge_weight(W, 0L, 1L)
strength(W, 1L)
total_weight(W)
update_edge_weight(W, 0L, 1L, 4.0)
```

## Removing edges

`remove_edge()` updates triangles, degrees and weights exactly. Union-Find
cannot split, so the next component query after a removal transparently
rebuilds it in O(n + m); everything else stays incremental.

```{r}
G2 <- stream_graph()
add_edge(G2, 0L, 1L); add_edge(G2, 1L, 2L)
remove_edge(G2, 1L, 2L)
n_components(G2)   # {0,1} plus the now-isolated node 2
```

Nodes may also be created explicitly and survive save/load as isolates:

```{r}
add_node(G2, 10L)
nodes(G2)
snapshot <- copy_graph(G2)
```

## Directed graphs

```{r}
D <- stream_graph(directed = TRUE)
add_edge(D, 0L, 1L); add_edge(D, 1L, 2L); add_edge(D, 2L, 0L); add_edge(D, 2L, 3L)

degree(D, 2L)        # out-degree
in_degree(D, 2L)
strong_component_ids(D)   # cycle {0,1,2} is one SCC, {3} another
n_components(D)           # weakly connected: all four nodes
```

Triangles are counted on the underlying undirected graph, `degree()` /
`neighbours()` are over out-edges, and components ignore direction (weak
components); use `strong_component_ids()` for direction-aware components.

## Centrality on demand

```{r}
P <- stream_graph()
for (i in 0:4) add_edge(P, i, i + 1L)

betweenness_approx(P, k = 1000L, seed = 42L)
pagerank(P)[1:6]
```

`betweenness_approx()` samples k random (s, t) pairs; increase k for
precision (it clamps to exact once k covers every pair). `normalise = FALSE`
returns raw dependency sums instead of [0, 1] scores.

## Interop and persistence

```{r, eval = requireNamespace("igraph", quietly = TRUE)}
ig <- as_igraph(G)      # igraph out
G3 <- from_igraph(ig)   # and back
```

```{r}
edge_list(G)[1:3, ]     # plain data.frame export

p <- tempfile(fileext = ".esg")
save_graph(G, p)
H <- load_graph(p)
n_edges(H) == n_edges(G)
```
